Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-08 text/ms.tex
Edit ms and supp to bring in new figs
A start

Black swans are statistically improbable events that nonetheless occur --- often with profound consequences. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). Black swans are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors . Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such. The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies --- particularly as the frequency and magnitude of climatic extremes increase over the next century.

@@

We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}. This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity. In any case,  for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}). TODO ADD ALTERNATIVE PRIOR NOTE.

TODO We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).

@@

We found strong, but rare, evidence for black-swan population dynamics. By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).

@@

TODO ADD TO THIS Our simulation testing show that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}, \ref{fig:sim-prob}) AND ``(as was also observed in simulation testing, Fig.~\ref{fig:sim-nu})''

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series length was the strongest covariate of observing black-swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps (Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with increased time series length was small ($0.\pHeavyNSixty$ vs.\ $0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

>>>>

Black swans are statistically improbable events that nonetheless occur --- often with profound consequences. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). Black swans are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors. Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such. The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies --- particularly as the frequency and magnitude of climatic extremes increase over the next century.

@@

We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}. This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity. In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to weaker and stronger priors (Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).

@@

We found strong, but rare, evidence for black-swan population dynamics. By defining black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our main Gompertz model found evidence for heavy tails most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, there was stronger evidence for black swans in insect populations than is visually apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median probability of heavy tails were insect orders --- however, there was considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}b).

@@

TODO sensitivity to priors Fig.~\ref{fig:alt-priors}.

TODO ADD TO THIS Our simulation testing show that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}) AND ``(as was also observed in simulation testing, Fig.~\ref{fig:sim-nu})''

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a). Longer time-series length was the strongest covariate of observing black-swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps. However, the absolute change in probability with increased time series length was small ($0.\pHeavyNSixty$ vs.\ $0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

### 2014-11-08 text/supp.tex
Edit ms and supp to bring in new figs
A start

\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population). We removed a limited number of populations from class Angiospermopsida and Bacillariophyceae to focus the taxonomy in our analysis. We also removed any populations with an \texttt{Unknown} taxonomic class.

@@

@@

We found that we could consistently and precisely recover median posterior estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$) (Fig.~\ref{fig:sim-gompertz} upper panels). At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were. At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Fig.~\ref{fig:sim-gompertz} lower panels).

@@

Our results show that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows). Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols). The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}. Fitting a model with correctly specified observation error made marginal improvements to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).

@@

To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig \ref{fig:order-estimates}), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor. (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.

@@

\singlespacing
\begin{footnotesize}
\begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{0.8cm}}
\caption{Populations with Pr$(\nu < 10) > 0.5$. TODO}\\
\toprule
\input{../analysis/cause-table.tex}
\label{tab:causes-supp}
\end{longtable}
\end{footnotesize}
\onehalfspacing

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/gomp-prior-comparison.pdf}
\caption{
Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. The base, weaker, and stronger priors on $\nu$ are illustrated in Fig.~\ref{fig:priors}. In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$. When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.
}
\label{fig:alt}
\end{center}
\end{figure}

>>>>

\item We removed GPDD Main IDs \texttt{20531} and \texttt{10139}, which we noticed were duplicates of \texttt{20579} (a heron population). \texttt{20579} contained additional years of data not present in \texttt{10139}. We removed a limited number of populations from class Angiospermopsida and Bacillariophyceae to focus the taxonomy in our analysis. We also removed any populations with an \texttt{Unknown} taxonomic class.

\item Finally, we removed populations with the following GPDD Main IDs, which we discovered data entry errors when verifying the populations with suspected black swans: \texttt{1207} because 1957 was entered as 2 but should have been 27 \citep{kendeigh1982}, \texttt{6531} because 1978 was entered as 7 should have been 47 \citep{minot1986}, and \texttt{6566} because the data did not match the graph \citep{heessen1996}.

@@

\section{Alternative priors}

TODO

Fig.~\ref{fig:priors}

Fig.~\ref{fig:alt-priors}

@@

We found that we could consistently and precisely recover median posterior estimates of $\nu$ near the true value of $\nu$ with large samples ($\ge 200$) (Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could still usually qualitatively distinguish heavy from not-heavy tails, but the model tended to underestimate how heavy the tails were. At the same time, at smaller sample sizes, the model tended to overestimate how large the scale parameter was (Fig.~\ref{fig:sim-nu} lower panels).

@@

TODO, rework this paragraph:
Our results show that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols in the top rows). Likewise, the other Gompertz parameters were estimated without any systematic bias (Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}, red and green symbols). The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness, overestimate the magnitude of process noise, somewhat overestimate $\lambda$, and overestimate density dependence (blue symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}). The overestimation of density dependence with observation error is a known phenomenon \citep{knape2012}. Fitting a model with correctly specified observation error made marginal improvements to model bias (purple symbols in Figs~\ref{fig:sim-gompertz} and \ref{fig:sim-gompertz-boxplots}).

@@

To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~\ref{fig:posteriors}b), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor. (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.

@@

\singlespacing
\begin{footnotesize}
\begin{longtable}{>{\RaggedRight}m{1.5cm}>{\RaggedRight}p{4.3cm}>{\RaggedRight}p{0.8cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.0cm}>{\RaggedRight}p{3.0cm}>{\RaggedRight}p{1.7cm}>{\RaggedRight}p{1.3cm}}
\caption{Populations with Pr$(\nu < 10) > 0.5$. TODO}\\
\toprule
\input{../analysis/cause-table.tex}
\label{tab:causes-supp}
\end{longtable}
\end{footnotesize}
\onehalfspacing

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/gomp-prior-comparison.pdf}
\caption{
Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. The base, weaker, and stronger priors on $\nu$ are illustrated in Fig.~\ref{fig:priors}. In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$. When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.
}
\label{fig:alt-priors}
\end{center}
\end{figure}
